Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-05-29 main_new.tex
Updates from Overleaf

\begin{abstract}
  We examine experiment design for informative sampling for
  characterizing oceanographic phenomena. Methods are developed
  for efficient spatial sampling using autonomous robotic platforms,
  where the goal is to map the regions of cold and fresh water near a
  river mouth. Specifically, we define a design criterion based on
  reducing uncertainty of excursion sets in the context of multivariate
  random fields. We develop analytical expressions for the expected
  variance reduction in excursion sets, and show how these can be
  applicable for static designs and sequential designs building on
  myopic and look-ahead planning. We use simulations to study the
  properties of the suggested methods. Results from field
  deployments with an autonomous underwater vehicle are also presented.

@@

Monitoring the worlds oceans has gained increased importance in the
light of the changing climate. While the focus has often been on
maga-fauna and the anthropogenic impact from micro-plastics to
pollution, biological oceanographers have focused intently on studying
micro-organisms at the base of the human food web. These organisms are
critically impacted by the changing dynamics in the upper water-column,
especially in coastal zones which are complex and often hard to observe
in space and time. By studying the bio-geochemical processes in the
upper water-column scientists can measure the impact of change, natural
or anthropomorphic, and provide an informed opinion to policy makers to
effect changes in preserving the environment. However, the challenge of 
% The pressure on marine resources is growing and increased accuracy,
% resolution, and persistent monitoring of the oceans is crucial for
% long-term sustainable management. A central problem for this type of
% environmental 
sensing in the small% in the large spatial extent
while studying the large spatio-temporal variation inherent to ocean
processes, prompts the need for effective means of sampling.
% allocation of sampling resources.
%\begin{em}
Sampling can, in this context, not simply be distributed uniformly
---along simple transects, grids, or ``lawn-mover" patterns--- but must
instead be prioritized to relevant regions to ensure it is
cost-effective while providing adequate coverage of the area of
scientific interest.
% save resources and avoid
% redundant data.
%\end{em}

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-centric methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to the
sensing the surface of the ocean and impacted by cloud cover. With the
advent of robust mobile robotic platforms \cite{Bellingham07}, more
effective and higher resolution methods are making significant
contributions to environmental monitoring and sampling. 
% A
% sustained focus on prioritized and efficient data collection strategies
% have therefore started to emerge. The advent of marine robotic
% platforms, especially
In particular Autonomous underwater vehicles (AUVs), have made
significant strides in advancing the state of sampling
% provided means to execute this prioritization through the capacity of
% autonomy and data-driven sampling, where data collection in principle
% can be optimized. These capabilities have made
and consequently have made robotics an integral part of ocean
observation in large part driven by using 
% of the emerging sensing practice for ocean science, allowing scientists
% to increase the observational efficiency and resolution beyond what was
% previously possible. But how should a robotic platform, such as an AUV,
% effectively prioritize and identify important regions for sampling? The
% answer to this question relates to 
statistical methods in design and data assimilation
\cite{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.% , and
% the application domain is clearly an arena where statisticians can
% contribute.
% There has recently been some statistical attention in oceanography:
Others include \cite{wikle2013modern} which % present an overview of approaches for modeling
% uncertainties in the ocean sciences with
focuses on hierarchical statistical models; \cite{sahu2008space} studies
spatio-temporal models for sea surface temperature and salinity data,
while
% which will also are our key variables of interest in the current paper.
\cite{mellucci2018oceanic} looks at the statistical prediction of
features using an underwater glider.
%From an oceanographic perspective, interesting regions are usually directly tied to a distinct phenomena that is of scientific interest. Each phenomenon can in turn be characterized by a set of process specific conditions expressed through different measures of key environmental variables, such as temperature or salinity. One such measure is the gradient, that can be associated with a number of important processes, such as the vertical location of the thermocline and pycnocline, location of upwelling systems, vertical mixing, eddies, fronts, and currents \cite{sverdrup2006}, as well as distribution, growth, and accumulation of biological activity \cite{SatOceanSoci00, Ryan2014}. These gradients create boundaries separating the ocean into process specific regions which are of profound interest to both identify and map effectively. Quantification of gradient features is therefore a much needed competence in robotic sampling.
The combination of statistical tools and robotic platforms is a natural
symbiosis which enables information-based sensing. Central to this is
the ability to model spatially correlated variables and provide formal
measures of uncertainty. By far the most practical models use Gaussian
processes (GPs), which will also be used in our formulation, in large
part because this approach enables evaluation and prioritization of
sampling efforts to be formalized into decision making systems on-board
a robotic platform.
%The imperfect understanding of causality coupled with the underlying fundamental uncertainty of the ocean implies that we need to obtain more measurements and use statistical models and methods to determine what and how phenomena form, to increase the quality of predictive statements.

In this work, we study excursion sets (ES) and excursion probabilities
(EP) for spatial GPs, and the goal is to design sampling strategies for
reducing the variance in the ES. The main contributions involve: i)
extending ES-based uncertainty criteria to multivariate random fields,
allowing the evaluation to include multiple environmental parameters
such as temperature and salinity, ii) develop methods for informative
environmental sensing and autonomous data collection, and iii)
demonstrating the approach in a real world scenario of sampling a river
plume using an AUV.

The remainder of the paper is organized as follows: Section \ref{sec:bg}
provides some background on ocean sampling, as also some notation for
ESs and EPs. Section \ref{sec:GP_EP} sets up the required assumptions of
GPs, including bi-variate random fields for temperature and salinity.
Section \ref{sec:sur} formulates the sampling design criteria for
variance reduction in ESs, with analytical derivation for multivariate
spatial processes. Section \ref{sec:heuristics} builds on these sampling
criteria to present heuristics for adaptive sampling. Section
\ref{sec:simulations} shows results and discusses properties of the
methods in simulation while Section \ref{sec:case_study} instantiates
the work in the form of characterizing a river plume. We conclude with
Section \ref{sec:concl_disc} with a summary and a discussion of future
work.

@@

River plumes belong to a class of ocean processes that are local and
usually act on smaller \emph{sub-mesoscale}, from 10 m -- 5 km where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detailed at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for fine
resolution is by direct observation. At larger \emph{mesoscale} ($>50$
km), both three-dimensional space and time dynamics are important and
can shift substantially, while in the case of river plumes one can often
limit scope to cover only lateral spatial elements ([north, east]
domain). For this case, time-effects can be regarded as static when
acquiring AUV data over a short time window and limited region. However,
the methodological framework can be extended to three-dimensional
spatial variability and temporal elements in a similar manner.

@@

% In the context illustrated by Figure \ref{fig:river4},
Our focus is on spatial separation of cold freshwater from the saline
waters of the fjord and to be able to classify these water masses. In
this context, ES variability is a useful sampling criterion for guiding
the observation effort of the AUV. A bivariate ES is defined via joint
variables being below thresholds, at individual locations, for the
process over the spatial domain $\mathcal{M}$. Denoting the threshold
for temperature by $T_a$ and the threshold for salinity by $T_b$, we use
the following definition;
 \begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_a(\bx) < T_a,\xi_b(\bx) < T_b\}.
 \end{equation}
The excursions could also be defined as larger than threshold, or within a boundary of limits for the two variables. Calculations and interpretations will be similar in these different cases.

The ES is defined by random indicator variables, and the associated EPs become 
\begin{equation}\label{eq:prob}
 p(\bx) = P(\xi_a(\bx) < T_a, \xi_b(\bx) < T_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
The actual calculation of the EP would require a model specification. As we demonstrate below, Gaussian modeling assumptions facilitate the computation. 
If the EPs in Eq. (\ref{eq:prob}) are close to $1$ or $0$, it is easy to classify water masses. It is more difficult if the probabilities are close to $0.5$. In this case it will likely be worthwhile gathering data to improve the classification skill.

@@

For implementation on the computer, we discretize the spatial domain $\mathcal{M}$ on a set of $n$ grid locations; $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$.
We let $\Delta$ be the area of each cell in the regular grid.
Using vector notation, we collect the temperature and salinity variables on the grid as
\begin{equation}
    \bxi=(\xi_a(\bx_1),\xi_b(\bx_1),\ldots,\xi_a(\bx_n),\xi_b(\bx_n))^t \sim \mbox{GP} (\bmu,\bSigma), \nonumber
\end{equation}
where the entries in length $2 n$ vector $\bmu$ hold the expected values of temperature and salinity at the grid locations, and the $2n \times 2n$ matrix $\bSigma$ holds all the covariances between temperature and salinity variables at all grid locations. 
 The joint probability density function is then
\begin{equation}\label{prior}
\pi(\bxi) = N(\bxi;\bmu,\bSigma) = \frac{1}{(2\pi)^{\frac{n}{2}}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi - \bmu)}.
\end{equation}

@@

Denote the AUV data at $m$ locations by $\by=(y_{a,1},y_{b,1},\ldots,y_{a,m},y_{b,m})^t$, representing the temperature and salinity measurements gathered along a survey line.
Restricting attention to the grid representation, the likelihood model is described by 
\begin{equation}\label{likelihood}
\by | \bxi \sim GP(\by; \bA \bxi,\bR),
\end{equation}
where $\bA$ is an $2m \times 2n$ matrix holding $1$ entries at the measurement indexes on the grid, and otherwise $0$ entries. The size $2m \times 2m$ covariance matrix $\bR$ is diagonal with entries $r_a$ and $r_b$ which are the measurement error variances of temperature and salinity observations.

The updated distribution for temperature and salinity variables on the spatial grid, given survey data $\by$, is Gaussian with mean and covariance
\begin{eqnarray}\label{gp_upd}
\bm &=& \bmu+\bSigma \bA^t (\bA \Sigma \bA^t+\bR)^{-1}(\by-\bA \bmu),  \\
\bS &=& \bSigma - \bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA \bSigma.\nonumber
\end{eqnarray}
The updated bivariate distribution at a grid location $\bx$ is then 
\begin{equation}\label{gp_hat}
\begin{bmatrix}
\xi_a(\bx) \\
\xi_b(\bx)
\end{bmatrix}
 |\by
 \sim N \left( 
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries extracted from the conditional mean and covariance expressions in Eq. (\ref{gp_upd}).

The conditional mean $\bm$ will be important in the following derivation in Section \ref{sec:sur}. This mean is a linear function of the data $\by$, and since the marginal distribution of the data is $\by \sim N( \bA \bmu,\bA \bSigma \bA^t+\bR)$, we get
\begin{equation}\label{distxi}
\bm \sim N(\bmu , \bPsi), \hspace{3mm} \bPsi=\bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA \bSigma.  
\end{equation}
When considering only grid location $\bx$, we can extract elements from the mean vector and covariance matrix to get the bivariate distribution as follows 
\begin{equation}\label{dist_mxi}
\bm(\bx) \sim N 
\left( 
\begin{bmatrix}
\mu_{a}(\bx) \\
\mu_{b}(\bx)
\end{bmatrix},
\begin{bmatrix}
\psi^2_{a}(\bx) & \psi_{a,b}(\bx)\\
\psi_{a,b}(\bx) & \psi^2_{b}(\bx)
\end{bmatrix}
\right).
\end{equation}

@@

Based on the Gaussian modeling assumptions, the EPs in Eq. (\ref{eq:prob}) can be computed at any location $\bx$. Without additional data, we have
\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_a(\bx) < T_a, \xi_b(\bx) < T_b) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} T_a\\
T_b
\end{bmatrix};
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx)\sigma_{b}(\bx) \gamma  \\
\sigma_{a}(\bx)\sigma_{b}(\bx) \gamma  & \sigma_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution function, and the mean and covariance entries are defined in Eq. (\ref{gp_bar}).

When data $\by$ is available, we similarly get
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_a(\bx) < T_a, \xi_b(\bx) < T_b |\by) 
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} T_a\\
T_b
\end{bmatrix};
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}).

@@

\subsection{Variance in excursion set}

@@

Consider then a measure as follows:
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx \approx \sum_{x \in \mathcal{M}_g} p(\bx) \left(1-p(\bx)\right) \Delta.
\end{equation}
The measure is dominated by location having probabilities close to $0.5$.

@@

To evaluate AUV data sampling paths, it is hence natural to evaluate the expected variance reduction in the ES  \citep{chevalier2014fast}, as provided by the survey data. In this way, we take the expectation with respect to the planned measurements $\by$:
\begin{eqnarray}\label{sur}
    V_{\mbox{upd}} &=& \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx \\
    & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} \Delta, \nonumber \\
    p(\bx;\by) &=& P(\xi_a(\bx) < T_a, \xi_b(\bx) < T_b |\by), \nonumber
\end{eqnarray}
where the conditional probability is defined in Eq. (\ref{eq:prob_mv}).

We will next derive closed form solutions to the expectation in (\ref{sur}). A critical
element in the derivation is the GP model, and the linear dependence on data $\by$, which means that we have a closed form Gaussian distribution for the conditional mean in Eq. (\ref{dist_mxi}). This means that the high-dimensional inner integral in Eq. (\ref{sur}) reduces to a univariate integral \citep{bhattacharjya2013value, chevalier2014fast}.

@@

The first key result is that the expectation only depends on the data through the linear combination $\bm=(m_{a},m_{b})^t$. 
To see this we standardize the two variables i.e.
$Z_a=\frac{\xi_a-m_{a}}{s_{a}}$,
$Z_b=\frac{\xi_b-m_{b}}{s_{b}}$; 
then
\begin{eqnarray}
   P(\xi_a < T_a, \xi_b < T_b|\by) &=& P \left( Z_a <\frac{T_a-m_{a}}{s_{a}}, Z_b <\frac{T_b-m_{b}}{s_{b}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{T_a-m_{a}}{s_{a}}\\
\frac{T_b-m_{b}}{s_{b}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_a < T_a, \xi_b < T_b|\bm),
\end{eqnarray}
where $\eta_{a,b} =s_{a,b}/(s_{a} s_{b})$ is defined in Eq. (\ref{eq:prob_mv}).
Since the expression only depends on $\by$ via $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$ (which will vary for each integrand term, because the locations differ).

The expected variance can now be written as
\begin{eqnarray}\label{eq:var2}
E_{\by}(p[1-p]) &=& E_{\bm}(p[1-p]), \hspace{3mm} p=P(\xi_a < T_a, \xi_b < T_b|\bm) \nonumber \\
E_{\bm}(p[1-p]) & = & \int p[1-p] p(\bm) d\bm, \nonumber \\
 &=& \int P(\xi_a < T_a, \xi_b < T_b|\bm)  p(\bm) d\bm \nonumber  \\
&-& \int P(\xi_a < T_a, \xi_b < T_b|\bm) P(\xi_a < T_a, \xi_b < T_b|\bm) p(\bm) d\bm, 
\end{eqnarray}
where the integral is split in two components.
We will next extend results of \cite{chevalier2014fast} to solve for these two parts.

>>>>

\begin{abstract}
  We examine experiment design for informative sampling for
  characterizing oceanographic phenomena. Methods are developed
  for efficient spatial sampling using autonomous robotic platforms,
  where the goal is to map the regions of cold and fresh water near a
  river mouth. Specifically, we define a design criterion based on
  reducing uncertainty in excursion sets of multivariate
  random fields. We develop analytical expressions for the expected
  variance reduction in excursion set indicators, and show how these can be
  applicable for static designs and sequential designs building on
  myopic and look-ahead planning. We use simulations to study the
  properties of the suggested methods. Results from field
  deployments with an autonomous underwater vehicle are also presented.

@@

Monitoring the worlds oceans has gained increased importance in the
light of the changing climate. While the focus has often been on
maga-fauna and the anthropogenic impact from micro-plastics to
pollution, biological oceanographers have focused intently on studying
micro-organisms at the base of the human food web. These organisms are
critically impacted by the changing dynamics in the upper water-column,
especially in coastal zones which are complex and often hard to observe
in space and time. By studying the bio-geochemical processes in the
upper water-column scientists can measure the impact of change, natural
or anthropomorphic, and provide an informed opinion to policy makers to
effect changes in preserving the environment. However, the challenge of 
% The pressure on marine resources is growing and increased accuracy,
% resolution, and persistent monitoring of the oceans is crucial for
% long-term sustainable management. A central problem for this type of
% environmental 
sensing in the small% in the large spatial extent
while studying the large spatio-temporal variation inherent to ocean
processes, prompts the need for effective means of sampling.
% allocation of sampling resources.
%\begin{em}
Sampling can, in this context, not simply be distributed evenly
---along simple transects, grids, or ``lawn-mover" patterns--- but must
instead be prioritized to relevant regions to ensure it is
cost-effective while providing adequate coverage of the area of
scientific interest.
% save resources and avoid
% redundant data.
%\end{em}

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-centric methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to the
sensing the surface of the ocean and impacted by cloud cover. With the
advent of robust mobile robotic platforms \cite{Bellingham07}, more
effective and higher resolution methods are making significant
contributions to environmental monitoring and sampling. 
% A
% sustained focus on prioritized and efficient data collection strategies
% have therefore started to emerge. The advent of marine robotic
% platforms, especially
In particular Autonomous underwater vehicles (AUVs), have made
significant strides in advancing the state of sampling
% provided means to execute this prioritization through the capacity of
% autonomy and data-driven sampling, where data collection in principle
% can be optimized. These capabilities have made
and consequently have made robotics an integral part of ocean
observation in large part driven by using 
% of the emerging sensing practice for ocean science, allowing scientists
% to increase the observational efficiency and resolution beyond what was
% previously possible. But how should a robotic platform, such as an AUV,
% effectively prioritize and identify important regions for sampling? The
% answer to this question relates to 
statistical methods in design and data assimilation
\cite{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.% , and
% the application domain is clearly an arena where statisticians can
% contribute.
% There has recently been some statistical attention in oceanography:
Others include \cite{wikle2013modern} who % present an overview of approaches for modeling
% uncertainties in the ocean sciences with
focus on hierarchical statistical models; \cite{sahu2008space} study
spatio-temporal models for sea surface temperature and salinity data,
while
% which will also are our key variables of interest in the current paper.
\cite{mellucci2018oceanic} look at the statistical prediction of
features using an underwater glider.
%From an oceanographic perspective, interesting regions are usually directly tied to a distinct phenomena that is of scientific interest. Each phenomenon can in turn be characterized by a set of process specific conditions expressed through different measures of key environmental variables, such as temperature or salinity. One such measure is the gradient, that can be associated with a number of important processes, such as the vertical location of the thermocline and pycnocline, location of upwelling systems, vertical mixing, eddies, fronts, and currents \cite{sverdrup2006}, as well as distribution, growth, and accumulation of biological activity \cite{SatOceanSoci00, Ryan2014}. These gradients create boundaries separating the ocean into process specific regions which are of profound interest to both identify and map effectively. Quantification of gradient features is therefore a much needed competence in robotic sampling.
The combination of statistical tools and robotic platforms is a natural
symbiosis which enables information-based sensing. Central to this is
the ability to model spatially correlated variables and provide formal
measures of uncertainty. By far the most practical models use Gaussian
processes (GPs), which will also be used in our formulation, in large
part because this approach allows efficient implementation and evaluation, enabling prioritization of
sampling efforts to be formalized into decision making systems on-board
a robotic platform.
%The imperfect understanding of causality coupled with the underlying fundamental uncertainty of the ocean implies that we need to obtain more measurements and use statistical models and methods to determine what and how phenomena form, to increase the quality of predictive statements.

In this work, we study excursion sets (ES) and excursion probabilities
(EP) for spatial GPs, and the goal is to design sampling strategies for
reducing the variance in the ES indicators. The main contributions involve: i)
extending ES-based uncertainty criteria to multivariate random fields,
allowing the evaluation to include multiple environmental parameters
such as temperature and salinity, ii) develop methods for informative
environmental sensing and autonomous data collection, and iii)
demonstrating the approach in a real world scenario of sampling a river
plume using an AUV.

The remainder of the paper is organized as follows: Section \ref{sec:bg}
provides some background on ocean sampling, as also some notation for
ESs and EPs. Section \ref{sec:GP_EP} sets up the required assumptions of
GPs, including bi-variate random fields for temperature and salinity.
Section \ref{sec:sur} formulates the choice of sampling design criteria for
variance reduction in ES indicators, with analytical derivation for multivariate
spatial processes. Section \ref{sec:heuristics} builds on these sampling
criteria to present heuristics for adaptive sampling. Section
\ref{sec:simulations} shows results and discusses properties of the
methods in simulation while Section \ref{sec:case_study} instantiates
the work in the form of characterizing a river plume. We conclude with
Section \ref{sec:concl_disc} with a summary and a discussion of future
work.

@@

River plumes belong to a class of ocean processes that are local and
usually act on smaller \emph{sub-mesoscale}, from 10 m -- 5 km where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detailed at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for fine
resolution is by direct observation. At larger \emph{mesoscale} ($>50$
km), both three-dimensional space and time dynamics are important and
can shift substantially, while in the case of river plumes one can often
limit scope to cover only lateral spatial elements ([north, east]
domain). Morover, for this case, time-effects can be regarded as static when
acquiring AUV data over a short time window and limited region. However,
the methodological framework can be extended to higher dimensional situations in a similar manner.

@@

% In the context illustrated by Figure \ref{fig:river4},
Our focus is on spatial separation of cold freshwater from the saline
waters of the fjord and to be able to classify these water masses. In
this context, ES variability is a useful sampling criterion for guiding
the observation effort of the AUV. A bivariate ES is defined via joint
variables being below thresholds, at individual locations, for the
process over the spatial domain $\mathcal{M}$. Denoting the threshold
for temperature by $T_a$ and the threshold for salinity by $T_b$, we use
the following definition;
 \begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_a(\bx) \leq T_a,\xi_b(\bx) \leq T_b\}.
 \end{equation}
The excursions could also be defined as larger than threshold, or within a boundary of limits for the two variables. Calculations and interpretations will be similar in these different cases.

The ES is defined by random indicator variables, and the associated EPs become 
\begin{equation}\label{eq:prob}
 p(\bx) = P(\xi_a(\bx) \leq T_a, \xi_b(\bx) \leq T_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
The actual calculation of the EP would require a model specification. As we demonstrate below, Gaussian modeling assumptions facilitate the computation. 
If the EPs in Eq. (\ref{eq:prob}) are close to $1$ or $0$, it is easy to classify water masses. It is more difficult if the probabilities are close to $0.5$. In this case it will likely be worthwhile gathering data to improve the classification skill.

@@

For implementation on the computer, we discretize the spatial domain $\mathcal{M}$ on a set of $n$ grid locations; $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$.
We let $\Delta$ be the area of each cell in the regular grid.
Using vector notation, we collect the temperature and salinity variables on the grid as
\begin{eqnarray}
    \bxi &=& (\xi_a(\bx_1),\xi_b(\bx_1),\ldots,\xi_a(\bx_n),\xi_b(\bx_n))^t, \nonumber \\
    \pi(\bxi) &=& N(\bmu,\bSigma) = \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi - \bmu)},
\end{eqnarray}
where $\pi(\bxi)$ is the joint density function of $\bxi$, entries in length $2 n$ vector $\bmu$ hold the expected values of temperature and salinity at the grid locations, and the $2n \times 2n$ matrix $\bSigma$ holds all the variance-covariances of temperature and salinity variables at all grid locations.

@@

Denote the AUV data at $m$ locations by $\by=(y_{a,1},y_{b,1},\ldots,y_{a,m},y_{b,m})^t$, representing the temperature and salinity measurements gathered along a survey line.
Restricting attention to the grid representation, the likelihood model is described by 
\begin{equation}\label{likelihood}
\pi(\by | \bxi)=N(\bA \bxi,\bR)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bA\bxi)^T\bR^{-1}(\by-\bA\bxi)},
\end{equation}
where $\bA$ is an $2m \times 2n$ matrix holding $1$ entries at the measurement indexes on the grid, and otherwise $0$ entries. The size $2m \times 2m$ covariance matrix $\bR$ is diagonal with entries $r_a$ and $r_b$ which are the measurement error variances of temperature and salinity observations.

The updated distribution for temperature and salinity variables on the spatial grid, given survey data $\by$, is Gaussian with mean and covariance
\begin{eqnarray}\label{gp_upd}
\bm &=& \bmu+\bSigma \bA^t (\bA \Sigma \bA^t+\bR)^{-1}(\by-\bA \bmu),  \\
\bS &=& \bSigma - \bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA \bSigma.\nonumber
\end{eqnarray}
The updated bivariate distribution at a grid location $\bx$ is then 
\begin{equation}\label{gp_hat}
\pi 
\left(
\begin{bmatrix}
\xi_a(\bx) \\
\xi_b(\bx)
\end{bmatrix}
 |\by
 \right)
 = N \left( 
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries extracted from the conditional mean and covariance expressions in Eq. (\ref{gp_upd}).

The conditional mean $\bm$ will be important in the following derivation in Section \ref{sec:sur}. This mean is a linear function of the data $\by$, and since the marginal distribution of the data is $\pi (\by)= N( \bA \bmu,\bA \bSigma \bA^t+\bR)$, we get
\begin{equation}\label{distxi}
\pi(\bm)= N(\bmu , \bPsi), \hspace{3mm} \bPsi=\bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA \bSigma.  
\end{equation}
When considering only grid location $\bx$, we can extract elements from the mean vector and covariance matrix to get the bivariate distribution as follows 
\begin{equation}\label{dist_mxi}
\pi \left( \bm(\bx) \right) 
= N 
\left( 
\begin{bmatrix}
\mu_{a}(\bx) \\
\mu_{b}(\bx)
\end{bmatrix},
\begin{bmatrix}
\psi^2_{a}(\bx) & \psi_{a,b}(\bx)\\
\psi_{a,b}(\bx) & \psi^2_{b}(\bx)
\end{bmatrix}
\right).
\end{equation}

@@

Based on the Gaussian modeling assumptions, the EPs in Eq. (\ref{eq:prob}) can be computed at any location $\bx$. Without additional data, we have
\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_a(\bx) \leq T_a, \xi_b(\bx) \leq T_b) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} T_a\\
T_b
\end{bmatrix};
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx)\sigma_{b}(\bx) \gamma  \\
\sigma_{a}(\bx)\sigma_{b}(\bx) \gamma  & \sigma_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution function, and the mean and covariance entries are defined in Eq. (\ref{gp_bar}).

When data $\by$ is available, we similarly get
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_a(\bx) \leq T_a, \xi_b(\bx) \leq T_b |\by) 
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} T_a\\
T_b
\end{bmatrix};
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}).

@@

\subsection{Variance in excursion set indicators}

@@

Consider then an integrated measure as follows:
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx \approx \sum_{x \in \mathcal{M}_g} p(\bx) \left(1-p(\bx)\right) \Delta.
\end{equation}
The measure is dominated by location having probabilities close to $0.5$. 
(Note that \cite{Bect2019} call this the integrated Bernoulli variance.)

@@

To evaluate AUV data sampling paths, it is hence natural to evaluate the expected variance reduction in the ES  \citep{chevalier2014fast}, as provided by the survey data. In this way, we take the expectation with respect to the planned measurements $\by$:
\begin{eqnarray}\label{sur}
    V_{\mbox{upd}} &=& \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx \\
    & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} \Delta, \nonumber \\
    p(\bx;\by) &=& P(\xi_a(\bx) \leq T_a, \xi_b(\bx) \leq T_b |\by), \nonumber
\end{eqnarray}
where the conditional probability is defined in Eq. (\ref{eq:prob_mv}).

We will next derive closed form solutions to the expectation in (\ref{sur}). A critical
element in the derivation is the GP model, and the linear dependence on data $\by$, which means that we have a closed form Gaussian distribution for the conditional mean in Eq. (\ref{dist_mxi}). This means that the high-dimensional inner integral in Eq. (\ref{sur}) reduces to a bivariate integral \citep{bhattacharjya2013value, chevalier2014fast}. The end result then relies on calls to the multivariate Gaussian cumulative distribution function for calculation of orthant probabilities.

@@

The first key result is that the expectation only depends on the data through the linear combination $\bm=(m_{a},m_{b})^t$. 
To see this we standardize the two variables i.e.
$Z_a=\frac{\xi_a-m_{a}}{s_{a}}$,
$Z_b=\frac{\xi_b-m_{b}}{s_{b}}$; 
then
\begin{eqnarray}
   P(\xi_a \leq T_a, \xi_b \leq T_b|\by) &=& P \left( Z_a \leq\frac{T_a-m_{a}}{s_{a}}, Z_b \leq\frac{T_b-m_{b}}{s_{b}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{T_a-m_{a}}{s_{a}}\\
\frac{T_b-m_{b}}{s_{b}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_a \leq T_a, \xi_b \leq T_b|\bm),
\end{eqnarray}
where $\eta_{a,b} =s_{a,b}/(s_{a} s_{b})$ is defined in Eq. (\ref{eq:prob_mv}).
Since the expression only depends on $\by$ via $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$ (which will vary for each integrand term, because the locations differ).
